### 2018-02-01 submission/manuscript.tex
run spell-check

\textbf{Methods.} DNA from Fecal samples (n = 4) were extracted using a
single PowerMag DNA extraction kit with a 10 minute bead beating step
and amplified at 15, 20, 25, 30, and 35 cycles using Accuprime, Kappa,
Phusion, Platinum, or Q5 HiFi DNA polymerase. Mock communities
(technical replicates n = 4) consisting of previously isolated whole
genomes of 8 different bacteria were also amplified using the same PCR
amplification approach. First, the number of OTUs (Operational Taxonomic
Units) was examined for both fecal samples and mock communites. Next,
Bray-Curtis index, the error rate, sequence error prevalence, and
chimera prevalence were assessed based on cycle number and HiFi DNA
polymerase. Finally, the chimera prevalence correlation with number of
OTUs was assessed for both cycle number and HiFi DNA polymerase
dependent differences.

@@

Over recent years there has been an increasing focus on standardizing
methodological approaches in microbiota research (Kim et al., 2017;
Hugerth \& Andersson, 2017). In particular, investigating ways that 16S
rRNA gene surveys can be made more replicatable has been a predominant
driver of this standardization push (Lauber et al., 2010; Salter et al.,
2014; Song et al., 2016; Gohl et al., 2016). Although 16S rRNA gene
sequencing has been much maligned for introduced bias, many of these
same considerations also affect metagenomic sequencing (Nayfach \&
Pollard, 2016; Costea et al., 2017). Between the two approaches similar
bias considerations include, but are not limited to, preservation media,
storage conditions, DNA extraction kit, PCR, and sequence library
preparation. Thus, for these overlapping considerations biases
identified for 16S rRNA gene sequencing will also likely influence
metagenomic sequencing results.

@@

\textbf{\emph{PCR Protocol:}} The five different high fidelity (HiFi)
Taq DNA polymerase that were tested were AccuPrime\textsuperscript{TM}
(ThermoFisher, MA, USA), KAPA HIFI (Roche, IN, USA), Phusion
(ThermoFisher, MA, USA), Platinum (ThermoFisher, MA, USA), and Q5 (New
England Biolabs, MA, USA). The PCR cycle conditions for Platinum and
Accuprime followed a previously published protocol (Kozich et al., 2013)
(\url{https://github.com/SchlossLab/MiSeq_WetLab_SOP/blob/master/MiSeq_WetLab_SOP_v4.md}).
The HiFi DNA polymerase activation time was 2 minutes, unless a
different activation was specified. For Kappa and Q5, the protocol
previously published by Gohl and colleagues was used (Gohl et al.,
2016). For Phusion, the company defined conditions were used except for
extension time, where the Accuprime and Platinum settings were used.

@@

\textbf{\emph{Sequence Processing:}} The mothur software program was
utilized for all sequence processing steps (Schloss et al., 2009).
Generally, the protocol followed what has been previously publsihed
(Kozich et al., 2013) (\url{https://www.mothur.org/wiki/MiSeq_SOP}). Two
major differences from the stated protocol were the use of VSEARCH
instead of UCHIME for chimera detection and the use of the OptiClust
algorithmn instead of average neighbor for Operational Taxonomic Unit
(OTU) generation at 97\% similarity (Edgar et al., 2011; Rognes et al.,
2016; Westcott \& Schloss, 2017). Sequence error was determined using
the seq.error command on mock samples after chimera removal and
classification to the RDP to remove non-bacterial sequences (Schloss et
al., 2009; Cole et al., 2013; Rognes et al., 2016).

\textbf{\emph{Statistical Analysis:}} All analysis was done with the R
(v 3.4.3) software package (R Core Team, 2017). Data tranformation and
graphing was completed using the tidyverse package (v 1.2.1) and colors
selected using the viridis package (v 0.4.1) (Garnier, 2017; Wickham,
2017). Differences in the total number of OTUs were analyzed using an
ANOVA with a tukey post-hoc test. For the fecal samples the data was
normalized to each individual by cycle number to account for the
biological variation between people. Bray-Curtis matrices were generated
using mothur after 100 sub-samplings at 1000, 5000, 10000, and 15000.
The distance matrix data was analyzed using PERMANOVA with the vegan
package (v 2.4.5) (Oksanen et al., 2017) and kruskal-wallis tests within
R. For both error and chimera analysis, samples were tested using
Kruskal-Wallis with a Dunns post-hoc test. Where applicable correction
for multiple comparison utilized the Benjamini-Hochberg method
(Benjamini \& Hochberg, 1995).

\textbf{\emph{Analysis Workflow:}} The total number of OTUs after
sub-sampling was analyzed for both the fecal and mock community samples.
Cycle dependent affects on Bray-Curits indicies were next assessed for
the fecal samples looking at both overall cycle differences and within
individual differences for the previous cycle (e.g.~20 versus 25, 25
versus 30, etc.). Based on these observations we wanted to next analyze
potential reasons for these differences. First, analysis of general
sequence error rate, number of sequences with an error, and base
subsitution were assessed in the mock community for each DNA polymerase.
After assessing these errors, the total number of chimeras was
determined after sequence processing. For the community based measures,
the fecal samples were analyzed at 4 different sub-sampling levels
(1000, 5000, 10000, and 15000) while the mock community samples were
analysed at 3 levels (1000, 5000, 10000).

\textbf{\emph{Reproducible Methods:}} The code and analysis can be found
here \url{https://github.com/SchlossLab/Sze_PCRSeqEffects_XXXX_2017}.
The raw sequences can be found in the SRA at the following accesssion
number \textbf{need to upload still}.

@@

\textbf{\emph{Number of OTUs is Dependent on the HiFi DNA Polymerase
Used:}} In order to compare the number of OTUs across individuals a
Z-score normalization by individual, by cycle number, was employed on
the number of OTUs data. After normalization, we identified that there
was a HiFi DNA polymerase dependent difference that was consistent
across sub-sampling {[}Figure 1{]}. Lower cycle numbers (15-20) tended
to result in less differences between HiFI DNA polymerase while cycle
numbers of 25, 30, and 35 tended to have more distinct differences
{[}Figure 1{]}. All subsampling levels were significantly different only
at 35 cycles (P-value \textless{} 0.0001) {[}Table S1{]}. There were
differences in HiFi DNA polymerase at 25 and 30 cycles but the
sub-sampling depth had to be 5000 or higher (P-value \textless{} 0.05)
{[}Table S1{]}. Platinum HiFi DNA polymerase was the main driver of the
differences observed across all sub-sampling depths at 35 cycles, based
on a Tukey post-hoc test (P-value \textless{} 0.05) {[}Table S2{]}.

@@

\textbf{\emph{Minimal Bray-Curtis Differences are Detected by Cycle
Number:}} Overall, there was very little difference between each
respective 5-cycle increment (e.g.~15x vs 20x) for both fecal and mock
samples and this was consistent across the different sub-samplings used
{[}Figure 3{]}. There were only two differences between 5-cycle
increments that were identified. First, there were differences for the
same fecal sample between 20x vs.~25x that was independent of HiFi DNA
polymerase but dependent on sub-sampling depth (subsampled to 1000 =
0.51 (0.4 - 0.79) (median (25\% - 75\% quantile)), subsampled to 5000 =
0.43 (0.33 - 0.63), subsampled to 10000 = 0.4 (0.24 - 0.43)) {[}Figure
3A{]}. Second, for the mocks, where data is available, there were larger
difference between 20x and 25x (subsampled to 1000 = 0.88 (0.42 - 0.91))
{[}Figure 3B{]}. However, these differences between the next 5 cycle
increment do not persist once 25 cylces are reached {[}Figure 3{]}.

Using PERMANOVA to test for differences within HiFi DNA polymerase
groups based on cycle number, only Phusion had cycle dependent
differences at 1000 and 5000 sub-sampling depth (P-value = 0.03 and
0.01, respectively). Phusion was one of only two HiFi DNA polymerases
that managed to have samples for the 1000 sub-sampling depth at 15
cycles. Next, we assessed whether there were any major differences
between 5 cycle increments within each sample. We found that there was
no detectable difference in Bray-Curits index when comparing to the
previous 5 cycle increment (P-value \textgreater{} 0.05). However,
Phusion at 1000 sub-sampling depth had a P-value = 0.02 before multiple
comparison correction. It should be noted that at higher sub-sampling
depths these differences in Bray-Curits indicies disappear for both
differences in cycle number and within 5 cycle increments within an
individual.

@@

The total seqeunces with at least one error had multiple differences at
different cycle numbers and was mostly alleviated by the use of the
pre.cluster step {[}Figure S1{]}. Major differences before this
pre.cluster step were driven by large differences in
Accuprime\textsuperscript{TM} and Platinum versus the other HiFi DNA
polymerase tested {[}Figure S1 \& Table S7 \& S8{]}. Although
Accuprime\textsuperscript{TM} had the lowest per base error rate it had
the largest number of sequences with at least one error, regardless of
cycle number or sub-sampling depth {[}Figure S1{]}. However, this
increased number of sequences with an error can be drastically lowered
with existing bioinformatic approaches {[}Figure S1{]}. Investigation of
whether there were HiFi DNA polymerase dependent effects on base
subsitution found that there were generally no biases in the types of
subsitution made {[}Figure S2{]}.

@@

Although we did not observe strong differences in the Bray-Curtis index,
based on cycle number, the data suggests that there may be differences
between 15 and 20 cycles versus higher cycle numbers, such as 30x, that
are commonly used. Additionally, there was no difference within
individuals between corresponding 5 cycle increments (e.g.~15 to 20, 20
to 25, etc.). Conversely, we may not have enough samples to adequately
judge the true magnitude difference between 20x and 25x communities and
higher cycle numbers analyzed, even though there is a clear trend to
suggest that 20x is very different then 25x {[}Figure 3{]}. This
finding, in conjunction with the PERMANOVA results, suggest that cycle
number can change bacterial community calculations but that these
differences are minimal once 25 cylces are reached. Increasing the
sub-sampling depth, for some DNA polymerase, may reduce some of these
observed community differences at lower cycle numbers.

Increasing the cycle number also exacerbated chimera prevalence
differences between the different HiFi DNA polymerases {[}Figure 5{]}.
The chimera prevalence was strongly correlated with the number of OTUs
which is relied upon heavily for many community metric calculations.
However, Bray-Curtis analysis with PERMANOVA showed few differences
based on DNA polymerase. It is possible that many of the increased
number of OTUs are not highly abundant allowing the Bray-Curtis index to
be able to successfully downweight these respective OTUs (Minchin,
1987). The choice of downstream diversity metric could be an important
consideration in helping to mitigate these observed changes due to high
chimera prevalence in specific HiFi DNA polymerases (e.g.~Platinum).

@@

\textbf{Figure 2: Mock Sample Variability in Number of OTUs based on
HiFi DNA Polymerase.} A) Sub-sampled to 1000 reads. B) Sub-sampled to
5000 reads. C) Sub-sampled to 10000 reads. The dotted line represents
the number of OTUs genereated when the mock reference sequences are run
through the pipeline.

>>>>

\textbf{Methods.} DNA from Fecal samples (n = 4) were extracted using a
single PowerMag DNA extraction kit with a 10 minute bead beating step
and amplified at 15, 20, 25, 30, and 35 cycles using Accuprime, Kappa,
Phusion, Platinum, or Q5 HiFi DNA polymerase. Mock communities
(technical replicates n = 4) consisting of previously isolated whole
genomes of 8 different bacteria were also amplified using the same PCR
amplification approach. First, the number of OTUs (Operational Taxonomic
Units) was examined for both fecal samples and mock communities. Next,
Bray-Curtis index, the error rate, sequence error prevalence, and
chimera prevalence were assessed based on cycle number and HiFi DNA
polymerase. Finally, the chimera prevalence correlation with number of
OTUs was assessed for both cycle number and HiFi DNA polymerase
dependent differences.

@@

Recently there has been an increasing focus on standardizing
methodological approaches in microbiota research (Kim et al., 2017;
Hugerth \& Andersson, 2017). In particular, investigating ways that 16S
rRNA gene surveys can be made more reproducible has been a predominant
driver of this standardization push (Lauber et al., 2010; Salter et al.,
2014; Song et al., 2016; Gohl et al., 2016). Although 16S rRNA gene
sequencing has been much maligned for introduced bias, many of these
same considerations also affect metagenomic sequencing (Nayfach \&
Pollard, 2016; Costea et al., 2017). Between the two approaches similar
bias considerations include, but are not limited to, preservation media,
storage conditions, DNA extraction kit, PCR, and sequence library
preparation. Thus, for these overlapping considerations biases
identified for 16S rRNA gene sequencing will also likely influence
metagenomic sequencing results.

@@

\textbf{\emph{PCR Protocol:}} The five different high fidelity (HiFi)
DNA polymerase that were tested were AccuPrime\textsuperscript{TM}
(ThermoFisher, MA, USA), KAPA HIFI (Roche, IN, USA), Phusion
(ThermoFisher, MA, USA), Platinum (ThermoFisher, MA, USA), and Q5 (New
England Biolabs, MA, USA). The PCR cycle conditions for Platinum and
Accuprime followed a previously published protocol (Kozich et al., 2013)
(\url{https://github.com/SchlossLab/MiSeq_WetLab_SOP/blob/master/MiSeq_WetLab_SOP_v4.md}).
The HiFi DNA polymerase activation time was 2 minutes, unless a
different activation was specified. For Kappa and Q5, the protocol
previously published by Gohl and colleagues was used (Gohl et al.,
2016). For Phusion, the company defined conditions were used except for
extension time, where the Accuprime and Platinum settings were used.

@@

\textbf{\emph{Sequence Processing:}} The mothur software program was
utilized for all sequence processing steps (Schloss et al., 2009).
Generally, the protocol followed what has been previously published
(Kozich et al., 2013) (\url{https://www.mothur.org/wiki/MiSeq_SOP}). Two
major differences from the stated protocol were the use of VSEARCH
instead of UCHIME for chimera detection and the use of the OptiClust
algorithm instead of average neighbor for Operational Taxonomic Unit
(OTU) generation at 97\% similarity (Edgar et al., 2011; Rognes et al.,
2016; Westcott \& Schloss, 2017). Sequence error was determined using
the seq.error command on mock samples after chimera removal and
classification to the RDP to remove non-bacterial sequences (Schloss et
al., 2009; Cole et al., 2013; Rognes et al., 2016).

\textbf{\emph{Statistical Analysis:}} All analysis was done with the R
(v 3.4.3) software package (R Core Team, 2017). Data transformation and
graphing was completed using the tidyverse package (v 1.2.1) and colors
selected using the viridis package (v 0.4.1) (Garnier, 2017; Wickham,
2017). Differences in the total number of OTUs were analyzed using an
ANOVA with a tukey post-hoc test. For the fecal samples the data was
normalized to each individual by cycle number to account for the
biological variation between people. Bray-Curtis matrices were generated
using mothur after 100 sub-samplings at 1000, 5000, 10000, and 15000.
The distance matrix data was analyzed using PERMANOVA with the vegan
package (v 2.4.5) (Oksanen et al., 2017) and kruskal-wallis tests within
R. For both error and chimera analysis, samples were tested using
Kruskal-Wallis with a Dunns post-hoc test. Where applicable correction
for multiple comparison utilized the Benjamini-Hochberg method
(Benjamini \& Hochberg, 1995).

\textbf{\emph{Analysis Workflow:}} The total number of OTUs after
sub-sampling was analyzed for both the fecal and mock community samples.
Cycle dependent affects on Bray-Curtis indices were next assessed for
the fecal samples looking at both overall cycle differences and within
individual differences for the previous cycle (e.g.~20 versus 25, 25
versus 30, etc.). Based on these observations we wanted to next analyze
potential reasons for these differences. First, analysis of general
sequence error rate, number of sequences with an error, and base
substitution were assessed in the mock community for each DNA
polymerase. After assessing these errors, the total number of chimeras
was determined after sequence processing. For the community based
measures, the fecal samples were analyzed at 4 different sub-sampling
levels (1000, 5000, 10000, and 15000) while the mock community samples
were analysed at 3 levels (1000, 5000, 10000).

\textbf{\emph{Reproducible Methods:}} The code and analysis can be found
here \url{https://github.com/SchlossLab/Sze_PCRSeqEffects_XXXX_2017}.
The raw sequences can be found in the SRA at the following accession
number \textbf{need to upload still}.

@@

\textbf{\emph{Number of OTUs is Dependent on the HiFi DNA Polymerase
Used:}} In order to compare the number of OTUs across individuals a
Z-score normalization by individual, by cycle number, was employed on
the number of OTUs data. After normalization, we identified that there
was a HiFi DNA polymerase dependent difference that was consistent
across sub-sampling {[}Figure 1{]}. Lower cycle numbers (15-20) tended
to result in less differences between HiFI DNA polymerase while cycle
numbers of 25, 30, and 35 tended to have more distinct differences
{[}Figure 1{]}. All sub-sampling levels were significantly different
only at 35 cycles (P-value \textless{} 0.0001) {[}Table S1{]}. There
were differences in HiFi DNA polymerase at 25 and 30 cycles but the
sub-sampling depth had to be 5000 or higher (P-value \textless{} 0.05)
{[}Table S1{]}. Platinum HiFi DNA polymerase was the main driver of the
differences observed across all sub-sampling depths at 35 cycles, based
on a Tukey post-hoc test (P-value \textless{} 0.05) {[}Table S2{]}.

@@

\textbf{\emph{Minimal Bray-Curtis Differences are Detected by Cycle
Number:}} Overall, there was very little difference between each
respective 5-cycle increment (e.g.~15x vs 20x) for both fecal and mock
samples and this was consistent across the different sub-samplings used
{[}Figure 3{]}. There were only two differences between 5-cycle
increments that were identified. First, there were differences for the
same fecal sample between 20x vs.~25x that was independent of HiFi DNA
polymerase but dependent on sub-sampling depth (sub-sampled to 1000 =
0.51 (0.4 - 0.79) (median (25\% - 75\% quantile)), sub-sampled to 5000 =
0.43 (0.33 - 0.63), sub-sampled to 10000 = 0.4 (0.24 - 0.43)) {[}Figure
3A{]}. Second, for the mocks, where data is available, there were larger
difference between 20x and 25x (sub-sampled to 1000 = 0.88 (0.42 -
0.91)) {[}Figure 3B{]}. However, these differences between the next 5
cycle increment do not persist once 25 cycles are reached {[}Figure
3{]}.

Using PERMANOVA to test for differences within HiFi DNA polymerase
groups based on cycle number, only Phusion had cycle dependent
differences at 1000 and 5000 sub-sampling depth (P-value = 0.03 and
0.01, respectively). Phusion was one of only two HiFi DNA polymerases
that managed to have samples for the 1000 sub-sampling depth at 15
cycles. Next, we assessed whether there were any major differences
between 5 cycle increments within each sample. We found that there was
no detectable difference in Bray-Curtis index when comparing to the
previous 5 cycle increment (P-value \textgreater{} 0.05). However,
Phusion at 1000 sub-sampling depth had a P-value = 0.02 before multiple
comparison correction. It should be noted that at higher sub-sampling
depths these differences in Bray-Curtis indices disappear for both
differences in cycle number and within 5 cycle increments within an
individual.

@@

The total sequences with at least one error had multiple differences at
different cycle numbers and was mostly alleviated by the use of the
pre.cluster step {[}Figure S1{]}. Major differences before this
pre.cluster step were driven by large differences in
Accuprime\textsuperscript{TM} and Platinum versus the other HiFi DNA
polymerase tested {[}Figure S1 \& Table S7 \& S8{]}. Although
Accuprime\textsuperscript{TM} had the lowest per base error rate it had
the largest number of sequences with at least one error, regardless of
cycle number or sub-sampling depth {[}Figure S1{]}. However, this
increased number of sequences with an error can be drastically lowered
with existing bioinformatic approaches {[}Figure S1{]}. Investigation of
whether there were HiFi DNA polymerase dependent effects on base
substitution found that there were generally no biases in the types of
substitution made {[}Figure S2{]}.

@@

Although we did not observe strong differences in the Bray-Curtis index,
based on cycle number, the data suggests that there may be differences
between 15 and 20 cycles versus higher cycle numbers, such as 30x, that
are commonly used. Additionally, there was no difference within
individuals between corresponding 5 cycle increments (e.g.~15 to 20, 20
to 25, etc.). Conversely, we may not have enough samples to adequately
judge the true magnitude difference between 20x and 25x communities and
higher cycle numbers analyzed, even though there is a clear trend to
suggest that 20x is very different then 25x {[}Figure 3{]}. This
finding, in conjunction with the PERMANOVA results, suggest that cycle
number can change bacterial community calculations but that these
differences are minimal once 25 cycles are reached. Increasing the
sub-sampling depth, for some DNA polymerase, may reduce some of these
observed community differences at lower cycle numbers.

Increasing the cycle number also exacerbated chimera prevalence
differences between the different HiFi DNA polymerases {[}Figure 5{]}.
The chimera prevalence was strongly correlated with the number of OTUs
which is relied upon heavily for many community metric calculations.
However, Bray-Curtis analysis with PERMANOVA showed few differences
based on DNA polymerase. It is possible that many of the increased
number of OTUs are not highly abundant allowing the Bray-Curtis index to
be able to successfully down-weight these respective OTUs (Minchin,
1987). The choice of downstream diversity metric could be an important
consideration in helping to mitigate these observed changes due to high
chimera prevalence in specific HiFi DNA polymerases (e.g.~Platinum).

@@

\textbf{Figure 2: Mock Sample Variability in Number of OTUs based on
HiFi DNA Polymerase.} A) Sub-sampled to 1000 reads. B) Sub-sampled to
5000 reads. C) Sub-sampled to 10000 reads. The dotted line represents
the number of OTUs generated when the mock reference sequences are run
through the pipeline.
